Mega-analysis of the Interoceptive Accuracy Scale (IAS) Structure and its Subjective Correlates

### 2025-10-24 paper/1_Initial/manuscript.qmd
Update manuscript.qmd

<!-- Study 1 findings -->
Study 1 demonstrated that the IAS is best conceptualised as comprising tightly coupled item pairs, reflecting discrete bodily domains such as visceral sensations, musculoskeletal cues, and expulsion phenomena. The pairs emerged after several items (e.g., Tickle, Taste, Affective Touch) were removed due to their instability, redundancy, or poor associations with the other items. The resulting 14-item IAS retained the psychometric integrity of the original scale while improving interpretability and efficiency.

A short version of the IAS has been proposed before.  @lin2023 suggested a 12-item version by removing items with local dependency, high difficulty, and/or low discrimination. Their parallel analysis suggested unidimensionality, although parallel analysis is known to underestimate factor numbers in multidimensional data [@golino2017exploratory; @markos2024dimensionality].

In contrast, our study applied multiple methods including EGA, FA, and HCA yet the same pairs consistently emerged across datasets and demonstrated superior fit compared to alternative models, which challenges the assumption of unidimensionality of the IAS [@murphy2019; @brand2023]. Although the short version of the IAS by  @lin2023 favoured a one-factor solution, their 12-item IAS was derived by removing locally dependent items identified through Item Response Theory (IRT).

We found that these locally dependent items can instead be interpreted as meaningful pairs (e.g., Heart–Breath), reflecting distinct bodily domains. Because IRT assumes unidimensionality [@nguyen2014introduction], it may have treated these natural item clusters as statistical artefacts rather than as theoretically relevant structures within interoceptive accuracy.

@@

Additionally, specific item pairings revealed distinct patterns of association with other interoceptive measures and mental health indices that were not captured by the full IAS. For instance, while the full IAS showed no significant associations with Borderline Personality, the Hunger-Thirst pair exhibited negative correlations. These findings collectively suggest that item-level analyses may uncover more nuanced relationships than those identified by the overall scale. This pattern highlights that questionnaires like the IAS may capture multiple facets of interoception rather than a single underlying dimension. Consequently, adopting a domain-specific approach could enhance construct validity by better reflecting the heterogeneity of interoceptive experiences.

These findings are in line with inconsistencies in the literature concerning associations between the IAS and the BPQ Body Awareness. While some studies report significant negative associations [@lin2023], others have observed significant positive associations  [@brand2023; @campos2021; @gaggero2021; @todd2022], and still others none at all  [@murphy2019]. In our study the full IAS was also not associated with the BPQ Body Awareness.

This pattern is consistent with the idea that this subscale may capture maladaptive bodily experiences or autonomic dysfunction, rather than interoceptive accuracy per se. However, the Body Awareness subscale of the BPQ has also been criticized for its ambiguous instructions. It is often unclear whether respondents should interpret the items as assessing attentional focus on bodily sensations, perceived accuracy in detecting them, or simply the frequency of experiencing such signals. As a result, studies have reported that observed associations vary depending on how the BPQ Body Awareness is interpreted [@campos2021; @gabriele2022dissociations; @gaggero2021]. As such, it is unclear whether the lack of a relationship between the IAS and the BPQ Body Awareness in this study reflects a genuine divergence between the constructs measured by the two scales, or instead arises from ambiguities in how the BPQ Body Awareness is operationalized and interpreted across studies.

@@

<!-- IAS limitations and potential improvements -->
These two studies revealed three main points. First,  in the datasets that included analogue-style ratings, these scales captured greater variability, further enhancing the ability to detect individual differences. Second, a shortened, 14-item version of the IAS retained strong psychometric integrity while removing unstable or redundant items. Third, specific item pairings allowed more precise differentiation among individuals and often outperformed global scores in predictive value. Nevertheless, even this refined IAS has important limitations.

>>>>

<!-- Study 1 findings -->
A short version of the IAS was previously proposed by @lin2023, who suggested a 12-item version by removing items with local dependency, high difficulty, and/or low discrimination. Their parallel analysis supported a unidimensional structure, although this method tends to underestimate factor numbers in multidimensional data [@golino2017exploratory; @markos2024dimensionality]. Consequently, the 12-item IAS was interpreted as reflecting a single underlying factor of interoceptive accuracy.

In contrast, Study 1 applied multiple complementary methods, including EGA, FA, and HCA. Across datasets and methods, the same tightly coupled item pairs consistently emerged, representing distinct bodily domains such as visceral sensations, musculoskeletal cues, and expulsion phenomena. Several items (e.g., Tickle, Taste, Affective Touch) were removed due to instability, redundancy, or weak associations, resulting in a refined 14-item IAS that retained psychometric robustness while improving interpretability and efficiency.

These findings challenge the assumption of IAS unidimensionality [@murphy2019; @brand2023]. Whereas @lin2023 ’s IRT-based approach treated locally dependent items as artefacts to be removed, our results suggest that such dependencies may reflect theoretically meaningful pairs (e.g., Heart–Breath). Because IRT assumes unidimensionality [@nguyen2014introduction], it may have obscured these structured item clusters rather than as theoretically relevant structures within interoceptive accuracy.

@@

Additionally, specific item pairings revealed distinct patterns of association with other interoceptive measures and mental health indices that were not captured by the full IAS. For instance, while the full IAS showed no significant associations with Borderline Personality, the Hunger-Thirst pair exhibited negative correlations. These findings collectively suggest that item-level analyses may uncover more nuanced relationships than those identified by the overall scale. This pattern highlights that questionnaires like the IAS may capture multiple facets of interoception rather than a single underlying dimension.

These findings are in line with inconsistencies found in the literature concerning associations between the IAS and the BPQ Body Awareness. While some studies report significant negative associations [@lin2023], others have observed significant positive associations  [@brand2023; @campos2021; @gaggero2021; @todd2022], and others none at all [@murphy2019]. In our study the full IAS was also not associated with the BPQ Body Awareness. This pattern is consistent with the idea that this subscale may capture maladaptive bodily experiences or autonomic dysfunction, rather than interoceptive accuracy per se.

However, the Body Awareness subscale of the BPQ has also been criticized for its ambiguous instructions. It is often unclear whether respondents should interpret the items as assessing attentional focus on bodily sensations, perceived accuracy in detecting them, or simply the frequency of experiencing such signals. As a result, studies have reported that observed associations vary depending on how the BPQ Body Awareness is interpreted [@campos2021; @gabriele2022dissociations; @gaggero2021]. As such, it is unclear whether the lack of a relationship between the IAS and the BPQ Body Awareness in this study reflects a genuine divergence between the constructs measured by the two scales, or instead arises from ambiguities in how the BPQ Body Awareness is operationalized and interpreted across studies.

@@

<!-- IAS limitations and potential improvements -->
Nevertheless, these two studies revealed three main points. First,  in the datasets that included analogue-style ratings, these scales captured greater variability, further enhancing the ability to detect individual differences. Second, a shortened, 14-item version of the IAS retained strong psychometric integrity while removing unstable or redundant items. Third, specific item pairings allowed more precise differentiation among individuals and often outperformed global scores in predictive value. Nevertheless, even this refined IAS has important limitations.
